\chapter{Time series models}\label{chap8}
In this chapter, we provide a brief introduction to performing inference in time series models using a Bayesian framework. There is a large literature in time series in statistics and econometrics, and it would be impossible to present a good treatment in a few pages of an introductory book. However, there are excellent books in Bayesian inference in time series, see for instance, \cite{west2006bayesian,petris2009dynamic,pole2018applied}.

A time series is a sequence of observations collected in chronological order, allowing us to track how variables change over time. However, it also introduces technical challenges, as we must account for statistical features such as autocorrelation and stationarity. Since time series data is time-dependent, we adjust our notation. Specifically, we use $t$ and $T$ instead of $i$ and $N$ to explicitly indicate time.

Our starting point in this chapter is the \textit{state-space representation} of time series models. Much of the Bayesian inference literature in time series adopts this approach, as it allows dynamic systems to be modeled in a structured way. This representation provides modularity, flexibility, efficiency, and interpretability in complex models where the state evolves over time. It also enables the use of recursive estimation methods, such as the \textit{Kalman filter} for dynamic Gaussian linear models and the \textit{particle filter} (also known as \textit{sequential Monte Carlo}) for non-Gaussian and nonlinear state-space models. The latter method is especially useful for \textit{online} predictions or when there are data storage limitations. These inferential tools are based on the sequential updating process of Bayes' rule, where the posterior at time $t$ becomes the prior at time $t+1$ (see Equation \ref{equpdate}).

Remember that we can run our GUI typing

\begin{tcolorbox}[enhanced,width=4.67in,center upper,
	fontupper=\large\bfseries,drop shadow southwest,sharp corners]
	\textit{R code. How to display our graphical user interface}
	\begin{VF}
		\begin{lstlisting}[language=R]
		shiny::runGitHub("besmarter/BSTApp", launch.browser = T)\end{lstlisting}
	\end{VF}
\end{tcolorbox} 

in the \textbf{R} package console or any \textbf{R} code editor, and once our GUI is deployed, select \textit{Time series Models}. However, users should see Chapter \ref{chapGUI} for details.

\section{State-space representation}\label{sec81}
A \textit{state-space model} is composed by of an \textit{unobservable state vector}  $\bm{\beta}_t \in \mathbb{R}^K$, and an \textit{observed} measure $\bm{Y}_t \in \mathbb{R}^M$, $t=1,2,\dots$ such that (i) $\bm{\beta}_t$ is a \textit{Markov process}, this is, $\pi(\bm{\beta}_t|\bm{\beta}_{1:t-1})=\pi(\bm{\beta}_t|\bm{\beta}_{t-1})$, all the information regarding $\bm{\beta}_t$ based on all its history up to $t-1$ is carried by $\bm{\beta}_{t-1}$, and (ii) $\bm{Y}_t$ is independent of $\bm{Y}_s$ conditional on $\bm{\beta}_t$, $t\neq s$ \cite[Chap.~2]{petris2009dynamic}.

These assumptions imply that $\pi(\bm{\beta}_{0:t},\bm{Y}_{1:t})=\pi(\bm{\beta}_0)\prod_{s=1}^{t}\pi(\bm{\beta}_s|\bm{\beta}_{s-1})\pi(\bm{Y}_s|\bm{\beta}_s)$.\footnote{A \textit{state-space model} where the states are random variables taking discrete values is called \textit{hidden Markov model}.}

There are three key aspects of \textit{state-space models}: \textit{filtering}, \textit{smoothing}, and \textit{forecasting}. In \textit{filtering}, we aim to estimate the current state given observations up to time $t$, specifically obtaining the density $\pi(\bm{\beta}_{s}|\bm{y}_{1:t})$ for $s = t$. In \textit{smoothing}, we conduct a retrospective analysis of the system, obtaining $\pi(\bm{\beta}_{s}|\bm{y}_{1:t})$ for $s < t$. In \textit{forecasting}, we forecast future observations by first obtaining $\pi(\bm{\beta}_{s}|\bm{y}_{1:t})$ as an intermediate step to compute $\pi(\bm{Y}_{s}|\bm{y}_{1:t})$ for $s > t$. A valuable feature of these methods is that all these densities can be calculated recursively. \cite{petris2009dynamic} show the recursive equations in Propositions 2.1 (filtering), 2.3 (smoothing) and 2.5 (forecasting).

An important class of \textit{state-space models} is the \textit{Gaussian linear state-space model}, also know as, \textit{dynamic linear model}:
\begin{align*}
	\bm{Y}_t&=\bm{X}_t\bm{\beta}_t+\bm{\mu}_t& \text{(Observation equations)}\\
	\bm{\beta}_t&=\bm{G}_t\bm{\beta}_{t-1}+\bm{w}_t & \text{(States equations)},
\end{align*}
where $\bm{\beta}_0\sim N(\bm{b}_0,\bm{B}_0)$, $\bm{\mu}_t\sim N(\bm{0}, \bm{\Sigma}_t)$, $\bm{w}_t\sim N(\bm{0}, \bm{\Omega}_t)$, $\bm{\beta}_0$, $\bm{\mu}_t$ and $\bm{w}_t$ are independent, $\bm{X}_t$ and $\bm{G}_t$ are $M\times K$ and $K\times K$ known matrices. Observe that this assumptions implies that $\bm{Y}_t|\bm{\beta}_t\sim N(\bm{X}_t\bm{\beta}_t,\bm{\Sigma}_t)$, and $\bm{\beta}_t|\bm{\beta}_{t-1}\sim N(\bm{G}_t\bm{\beta}_{t-1},\bm{\Omega}_t)$.

Let $\bm{\beta}_{t-1}|\bm{y}_{1:t-1}\sim N(\bm{b}_{t-1},\bm{B}_{t-1})$, then, we can get the \textit{Kalman filter} by getting
\begin{enumerate}
	\item The one-step-ahead predictive distribution of $\bm{\beta}_t$ given $\bm{y}_{1:t-1}$ is $\bm{\beta}_t|\bm{y}_{1:t-1}\sim N(\bm{a}_t, \bm{R}_t)$, where $\bm{a}_t=\bm{G}_t\bm{b}_{t-1}$ and $\bm{R}_t=\bm{G}_t\bm{B}_{t-1}\bm{G}_t^{\top}+\bm{\Omega}_t$.
	\item  The one-step-ahead predictive distribution of $\bm{Y}_t$ given $\bm{y}_{1:t-1}$ is $\bm{Y}_t|\bm{y}_{1:t-1}\sim N(\bm{f}_t, \bm{Q}_t)$, where $\bm{f}_t=\bm{X}_t\bm{a}_t$ and $\bm{Q}_t=\bm{X}_t\bm{R}_t\bm{X}_t^{\top}+\bm{\Sigma}_t$.
	\item The distribution of the one-step-ahead prediction error  $\bm{e}_t=\bm{Y}_t-\mathbb{E}[\bm{Y}_t|\bm{y}_{1:t-1}]=\bm{Y}_t-\bm{f}_t$ is $N(\bm{0}, \bm{Q}_t)$ \cite[Chap.~6]{shumway2017time}. 
	\item The filtering distribution of $\bm{\beta}_t$ given $\bm{y}_{1:t}$ is $\bm{\beta}_t|\bm{y}_{1:t}\sim N(\bm{b}_t, \bm{B}_t)$, where $\bm{b}_t=\bm{a}_t+\bm{K}_t\bm{e}_t$, $\bm{K}_t=\bm{R}_t\bm{X}_t^{\top}\bm{Q}_t^{-1}$ is the \textit{Kalman gain}, and $\bm{B}_t=\bm{R}_t-\bm{R}_t\bm{X}_t^{\top}\bm{Q}_t^{-1}\bm{X}_t\bm{R}_t$.   
\end{enumerate}    

The formal proofs of these results can be found in \cite[Chap~2]{petris2009dynamic}. Just take into account that the logic of these results follow the results of the Seemingly unrelated regression (SUR) model in Section \ref{sec72} for a particular time period. In addition, we know that the posterior distribution using information up to $t-1$ becomes the prior in $t$ (see Equation \ref{equpdate}, $\pi(\bm{\theta}|\mathbf{y}_{1:t})\propto p(y_{t}|\mathbf{y}_{1:t-1},\bm{\theta})\times \pi(\bm{\theta}|\mathbf{y}_{1:t-1})$). This is the updating process from  $\bm{\beta}_t|\bm{y}_{1:t-1}\sim N(\bm{a}_t, \bm{R}_t)$ to $\bm{\beta}_t|\bm{y}_{1:t}\sim N(\bm{b}_t, \bm{B}_t)$. Moreover, the posterior mean and variance of the SUR model with independent conjugate priors for a particular time period can be written as $\bm{a}_{t}+\bm{R}_{t}\bm{X}_t^{\top}(\bm{X}_t\bm{R}_{t}\bm{X}_t^{\top}+ \bm{\Sigma}_t)^{-1}(\bm{y}_t-\bm{X}_t\bm{a}_{t})$ and $\bm{R}_{t}-\bm{R}_{t}\bm{X}_t^{\top}(\bm{X}_t\bm{R}_{t}\bm{X}_t^{\top}+\bm{\Sigma}_t)^{-1} \bm{X}_t\bm{R}_{t}^{\top}$, respectively. Let's see this, we know from Section \ref{sec72} that $\bm{B}_t=(\bm{R}_t^{-1}+\bm{X}_t^{\top}\bm{\Sigma}^{-1}\bm{X}_t)^{-1}$ and $\bm{\beta}_t=\bm{B}_t(\bm{R}_t^{-1}\bm{a}_t+\bm{X}_t^{\top}\bm{\Sigma}^{-1}\bm{y}_t)$. Thus, let's show that both conditional posterior distributions are the same. In particular, the posterior mean in the \textit{state-space representation} is $[\bm{I}_K-\bm{R}_{t}\bm{X}_t^{\top}(\bm{X}_t\bm{R}_{t}\bm{X}_t^{\top}+ \bm{\Sigma}_t)^{-1}\bm{X}_t]\bm{a}_{t}+\bm{R}_{t}\bm{X}_t^{\top}(\bm{X}_t\bm{R}_{t}\bm{X}_t^{\top}+ \bm{\Sigma}_t)^{-1}\bm{y}_t$, where 
\begin{align*}
	\bm{R}_{t}\bm{X}_t^{\top}(\bm{X}_t\bm{R}_{t}\bm{X}_t^{\top}+ \bm{\Sigma}_t)^{-1}
	&=\bm{R}_{t}\bm{X}_t^{\top}[\bm{\Sigma}_t^{-1}-\bm{\Sigma}_t^{-1}\bm{X}_t(\bm{R}_t^{-1}+\bm{X}_t^{\top}\bm{\Sigma}_t^{-1}\bm{X}_t)^{-1}\bm{X}_t^{\top}\bm{\Sigma}_t^{-1}]\\
	&=\bm{R}_{t}[\bm{I}_K-\bm{X}_t^{\top}\bm{\Sigma}_t^{-1}\bm{X}_t(\bm{R}_t^{-1}+\bm{X}_t^{\top}\bm{\Sigma}_t^{-1}\bm{X}_t)^{-1}]\bm{X}_t^{\top}\bm{\Sigma}_t^{-1}\\
	&=\bm{R}_{t}(\bm{I}_K-[\bm{I}_K-\bm{R}_t^{-1}(\bm{R}_t^{-1}+\bm{X}_t^{\top}\bm{\Sigma}_t^{-1}\bm{X}_t)^{-1}])\bm{X}_t^{\top}\bm{\Sigma}_t^{-1}\\
	&=(\bm{R}_t^{-1}+\bm{X}_t^{\top}\bm{\Sigma}_t^{-1}\bm{X}_t)^{-1}\bm{X}_t^{\top}\bm{\Sigma}_t^{-1},
\end{align*}
where the first equality uses the Woodbury matrix identity (matrix inversion lemma), and the third equality uses $\bm{D}(\bm{D}+\bm{E})^{-1}=\bm{I}-\bm{E}(\bm{D}+\bm{E})^{-1}$. 

Thus, $[\bm{I}_K-\bm{R}_{t}\bm{X}_t^{\top}(\bm{X}_t\bm{R}_{t}\bm{X}_t^{\top}+ \bm{\Sigma}_t)^{-1}\bm{X}_t]\bm{a}_{t}+\bm{R}_{t}\bm{X}_t^{\top}(\bm{X}_t\bm{R}_{t}\bm{X}_t^{\top}+ \bm{\Sigma}_t)^{-1}\bm{y}_t=[\bm{I}_K-(\bm{R}_t^{-1}+\bm{X}_t^{\top}\bm{\Sigma}_t^{-1}\bm{X}_t)^{-1}\bm{X}_t^{\top}\bm{\Sigma}_t^{-1}\bm{X}_t]\bm{a}_{t}+(\bm{R}_t^{-1}+\bm{X}_t^{\top}\bm{\Sigma}_t^{-1}\bm{X}_t)^{-1}\bm{X}_t^{\top}\bm{\Sigma}_t^{-1}\bm{y}_t=(\bm{R}_t^{-1}+\bm{X}_t^{\top}\bm{\Sigma}_t^{-1}\bm{X}_t)^{-1}\bm{R}_t^{-1}\bm{a}_{t}+(\bm{R}_t^{-1}+\bm{X}_t^{\top}\bm{\Sigma}_t^{-1}\bm{X}_t)^{-1}\bm{X}_t^{\top}\bm{\Sigma}_t^{-1}\bm{y}_t=(\bm{R}_t^{-1}+\bm{X}_t^{\top}\bm{\Sigma}_t^{-1}\bm{X}_t)^{-1}(\bm{R}_t^{-1}\bm{a}_{t}+\bm{X}_t^{\top}\bm{\Sigma}_t^{-1}\bm{y}_t)$. The second equality uses $\bm{I}-(\bm{D}+\bm{E})^{-1}\bm{D}=(\bm{D}+\bm{E})^{-1}\bm{E}$. 

The equality of variances of both approaches is as follows:
\begin{align*}
	Var[\bm{\beta}_t|\bm{y}_{1:t}]&
	= \bm{R}_{t}-\bm{R}_{t}\bm{X}_t^{\top}(\bm{X}_t\bm{R}_{t}\bm{X}_t^\top+\bm{\Sigma}_t)^{-1} \bm{X}_t\bm{R}_{t}\\
	&=\bm{R}_{t}-\bm{R}_{t}\bm{X}_t^{\top}(\bm{\Sigma}_t^{-1}- \bm{\Sigma}_t^{-1}\bm{X}_t(\bm{R}_{t}^{-1}+\bm{X}_t^{\top}\bm{\Sigma}_t^{-1}\bm{X}_t)^{-1}\bm{X}_t^{\top}\bm{\Sigma}_t^{-1})\bm{X}_t\bm{R}_{t}\\
	&=\bm{R}_{t}-\bm{R}_{t}\bm{X}_t^{\top}\bm{\Sigma}_t^{-1}\bm{X}_t\bm{R}_{t}+ \bm{R}_{t}\bm{X}_t^{\top}\bm{\Sigma}_t^{-1}\bm{X}_t(\bm{R}_{t}^{-1}+\bm{X}_t^{\top}\bm{\Sigma}_t^{-1}\bm{X}_t)^{-1}\bm{X}_t^{\top}\bm{\Sigma}_t^{-1}\bm{X}_t\bm{R}_{t}\\
	&=\bm{R}_{t}-\bm{R}_{t}\bm{X}_t^{\top}\bm{\Sigma}_t^{-1}\bm{X}_t\bm{R}_{t}+ \bm{R}_{t}\bm{X}_t^{\top}\bm{\Sigma}_t^{-1}\bm{X}_t[\bm{I}_K-(\bm{R}_{t}^{-1}+\bm{X}_t^{\top}\bm{\Sigma}_t^{-1}\bm{X}_t)^{-1}\bm{R}_{t}^{-1}]\bm{R}_{t}\\
	&=\bm{R}_{t}-\bm{R}_{t}\bm{X}_t^{\top}\bm{\Sigma}_t^{-1}\bm{X}_t(\bm{R}_{t}^{-1}+\bm{X}_t^{\top}\bm{\Sigma}_t^{-1}\bm{X}_t)^{-1}\\
	&=\bm{R}_t[\bm{I}_K-\bm{X}_t^{\top}\bm{\Sigma}_t^{-1}\bm{X}_t(\bm{R}_{t}^{-1}+\bm{X}_t^{\top}\bm{\Sigma}_t^{-1}\bm{X}_t)^{-1}]\\
	&=\bm{R}_{t}[\bm{I}_K-(\bm{I}_K-\bm{R}_{t}^{-1}(\bm{R}_{t}^{-1}+\bm{X}_t^{\top}\bm{\Sigma}_t^{-1}\bm{X}_t)^{-1})]\\
	&=(\bm{R}_{t}^{-1}+\bm{X}_t^{\top}\bm{\Sigma}_t^{-1}\bm{X}_t)^{-1},
\end{align*}
where the second equality uses the Woodbury matrix identity, the fourth equality uses $(\bm{D}+\bm{E})^{-1}\bm{D}=\bm{I}-(\bm{D}+\bm{E})^{-1}\bm{E}$, and the seventh equality uses $\bm{D}(\bm{D}+\bm{E})^{-1}=\bm{I}-\bm{E}(\bm{D}+\bm{E})^{-1}$.  

The \textit{Kalman filter} allows calculating recursively in a forward way $\pi(\bm{\beta}_t|\bm{y}_{1:t})$ from $\pi(\bm{\beta}_{t-1}|\bm{y}_{1:t-1})$ starting from $\pi(\bm{\beta}_0)$.

Let $\bm{\beta}_{t+1}|\bm{y}_{1:T}\sim N(\bm{s}_{t+1},\bm{S}_{t+1})$, then we can get the \textit{Kalman smoother} by $\bm{\beta}_{t}|\bm{y}_{1:T}\sim N(\bm{s}_{t},\bm{S}_{t})$, where $\bm{s}_t=\bm{b}_t+\bm{B}_t\bm{G}_{t+1}^{\top}\bm{R}_{t+1}^{-1}(\bm{s}_{t+1}-\bm{a}_{t+1})$ and $\bm{S}_t=\bm{B}_t-\bm{B}_t\bm{G}_{t+1}^{\top}\bm{R}_{t+1}^{-1}(\bm{R}_{t+1}-\bm{S}_{t+1})\bm{R}_{t+1}^{-1}\bm{G}_{t+1}\bm{B}_{t}$. The proof can be found in \cite[Chap~2]{petris2009dynamic}. 

Thus, we can calculate the \textit{Kalman smoother} starting from $t=T-1$, that is, $\bm{\beta}_{T}|\bm{y}_{1:T}\sim N(\bm{s}_{T},\bm{S}_{T})$. However, this is the filtering distribution at $T$, which means $\bm{s}_{T}=\bm{b}_{T}$ and $\bm{S}_{T}=\bm{B}_{T}$, and then, we should proceed recursively in a backward way.

Finally, the forecasting recursion in the \textit{dynamic linear model}, given $\bm{a}_t(0)=\bm{b}_t$ and $\bm{R}_t(0)=\bm{B}_t$, $h\geq 1$, is given by 
\begin{enumerate}
	\item The forecasting distribution of $\bm{\beta}_{t+h}|\bm{y}_{1:t}$ is $N(\bm{a}_t(h),\bm{R}_t(h))$, where $\bm{a}_t(h)=\bm{G}_{t+h}\bm{a}_{t}(h-1)$ and $\bm{R}_t(h)=\bm{G}_{t+h}\bm{R}_t(h-1)\bm{G}_{t+h}^{\top}+\bm{\Omega}_{t+h}$.
	\item The forecasting distribution $\bm{Y}_{t+h}|\bm{y}_{1:t}$ is $N(\bm{f}_t(h),\bm{Q}_t(h))$, where $\bm{f}_t(h)=\bm{X}_{t+h}\bm{a}_t(h)$ and $\bm{Q}_t(h)=\bm{X}_{t+h}\bm{R}_t(h)\bm{X}_{t+h}^{\top}+\bm{\Sigma}_{t+h}$.  
\end{enumerate}
The proof can be found in \cite[Chap~2]{petris2009dynamic}.

These recursive equations allows to perform probabilistic forecasting $h$-steps-ahead for the state and observation equations.

These results show how to use these recursive equations to perform filtering, smoothing and forecasting in \textit{dynamic linear models} (\textit{Gaussian linear state-space models}). Despite that these algorithms look simple, they suffer from numerical instability that lead to nonsymmetric and negative definite calculated covariance matrices. Thus, special care should be put when working with them.

\subsection{ARMA processes}

Let's see the \textit{state-space} representation of the $ARMA(p,q)$ model $y_t=\sum_{s=1}^{p}\phi_jy_{t-s}+\sum_{s=1}^{q}\theta_s e_{t-s}+e_t$. Setting $r=\max \left\{p,q+1\right\}$, $\phi_s=0$ for $s>p$ and $\theta_s=0$ for $s>q$, and defining the matrices $\bm{X}=[1 \ 0 \ \dots \ 0]$, $\bm{R}=[1 \ \psi_1 \ \dots \ \psi_{r-1}]^{\top}$, both are $r$-dimensional vectors, 
\begin{align*}
	\bm{G}=\begin{bmatrix}
		\phi_1 & 1 & 0 & \dots & 0\\
		\phi_2 & 0 & 1 & \dots & 0\\
		\vdots & \vdots & \ddots &  &\\
		\phi_{r-1} & 0 & 0 & \dots & 1\\
		\phi_r & 0 & 0 & \dots & 0\\
	\end{bmatrix} = \begin{bmatrix}
		\phi_1 & \vdots &  &  & \\
		\phi_2 & \vdots &  & \bm{I}_{r-1}  & \\
		\vdots & \vdots &  &  &\\
		\dots & \dots & \dots & \dots & \dots\\
		\phi_r & 0 & 0 & \dots & 0\\
	\end{bmatrix},
\end{align*} 
which is a $r\times r$ dimensional matrix, and give the \textit{state} vector $\bm{\beta}_t=[\beta_{1,t} \ \beta_{2,t} \ \dots \ \beta_{r,t}]^{\top}$, the $ARMA$ model has the following representation:
\begin{align*}
	y_t&=\bm{X}\bm{\beta}_t\\
	\bm{\beta}_t &= \bm{G}\bm{\beta}_{t-1}+\bm{R}e_{t}.
\end{align*}

This is a \textit{dynamic linear model} where $\bm{\Sigma}_t=0$, and $\bm{\Omega}_t=\sigma^2\bm{R}\bm{R}^{\top}$ (see \cite{petris2009dynamic,chib1994bayes}).

A nice advantage of the \textit{state-space} representation of the $ARMA$ model is that the evaluation of the likelihood can be performed efficiently using the recursive laws. Extensions to $ARIMA(p,d,q)$ models can be seen in \cite[Chap.~3]{petris2009dynamic}.\\

\textbf{Example: $AR(2)$ process}

Let's see the \textit{state-space} representation of a stationary $AR(2)$ process with intercept, that is, $y_t=\mu+\phi_1y_{t-1}+\phi_2y_{t-2}+e_t$, where $e_t\sim N(0,\sigma^2)$. Thus, $\mathbb{E}[y_t]=\frac{\mu}{1-\phi_1-\phi_2}$, and variance $Var[y_t]=\frac{\sigma^2(1-\phi_2)}{1-\phi_2-\phi_1^2-\phi_1^2\phi_2-\phi_2^2+\phi_2^3}$.

In addition, we can proof that setting $z_t=y_t-\bar{\mu}$, we have $z_t=\phi_1z_{t-1}+\phi_2z_{t-2}+e_t$ where $\mathbb{E}[z_t]=0$, and these are equivalent representations (see Exercise 1). Then, setting  $\bm{X}=[1 \ 0]$, $\bm{R}=[1 \ 0]^{\top}$, $\bm{G}=\begin{bmatrix}
	\phi_1 & 1\\
	\phi_2 & 0 \\
\end{bmatrix}$, $\bm{\beta}_t=[\beta_{1,t} \ \beta_{2,t}]^{\top}$, $\bm{\Sigma}_t=0$ and $\bm{\Omega}_t=\sigma^2$ we have
\begin{align*}
	z_t&=\bm{X}_t\bm{\beta}_t+\bm{\mu}_t& \text{(Observation equations)}\\
	\bm{\beta}_t&=\bm{G}_t\bm{\beta}_{t-1}+\bm{w}_t & \text{(States equations)}.
\end{align*}  


\section{Autoregressive models}\label{sec82}

Let's start with the linear Gaussian model with autorregresive errors,

\begin{align}
	y_t & = \bm{x}_t^{\top}\bm{\beta}+\mu_t\label{eq1}\\
	\phi(L)\mu_t & = e_t \label{eq2}, 
\end{align}
where $e_t \stackrel{iid}{\sim} N(0,\sigma^2)$, $\phi(L)=1-\phi_1L-\phi_2L^2-\dots-\phi_pL^p$ is a polynomial in the lag operator ($L$), where $Lz_t=z_{t-1}$, and in general, $L^rz_t=z_{t-r}$.

Thus, we see that stochastic error $\mu_t$ follows an \textit{autoregressive process of order $p$}, that is, $\mu_t\sim AR(p)$. It is standard practice to assume that $\mu_t$ is second-order stationary, this implies that the mean, variance and covariance of $\mu_t$ are constant. This implies that all roots of $\phi(L)$ lie outside the unit circle.

The likelihood function conditional on the first $p$ observations is
\begin{align*}
	p(y_{p+1},\dots,y_T|y_{p},\dots,y_1,\bm{\theta})&=\prod_{t=p+1}^{T}p(y_t|H_{t-1},\bm{\theta})\\
	&\propto \sigma^{-(T-p)}\exp\left\{-\frac{1}{2\sigma^2}\sum_{t=p+1}^T(y_t-\hat{y}_{t|t-1,\bm{\theta}})^2\right\},
\end{align*} 
where $H_{t-1}$ is the past history, $\bm{\theta}$ collects all parameters ($\bm{\theta}, \phi_1,\dots,\phi_p, \sigma^2$), and $\hat{y}_{t|t-1,\bm{\theta}}=(1-\phi(L))y_t+\phi(L)\bm{x}^{\top}\bm{\beta}$.

We can see that multiplying the first expression in Equation \ref{eq1} by $\phi(L)$, we can express the model as 
\begin{align}\label{eq3}
	y_t^*=\bm{x}_t^{*\top}\bm{\beta}+e_t
\end{align}
where $y_t^*=\phi(L)y_t$ and $\bm{x}_t^{*}=\phi(L)\bm{x}_t$.

Thus, collecting all observations $t=p+1,p+2,\dots,T$, we have $\bm{y}^*=\bm{X}^*\bm{\beta}+\bm{e}$, where $\bm{e}\sim N(\bm{0},\sigma^2\bm{I}_{T-p})$, $\bm{y}^*$ is a $T-p$ dimensional vector, and $\bm{X}^*$ is a $(T-p)\times K$ dimensional matrix.

Assuming that $\bm{\beta}|\sigma\sim N(\bm{\beta}_0,\sigma^2\bm{B}_0)$, $\sigma^2\sim IG(\alpha_0/2,\delta_0/2)$ and $\bm{\phi}\sim N(\phi_0,\bm{\Phi}_0)\mathbbm{1}[\bm{\phi}\in S_{\phi}]$, where $S_{\bm{\phi}}$ is the stationary region of $\bm{\phi}=[\phi_1 \ \dots \ \phi_p]^{\top}$. Then, Equation \ref{eq3} implies that $\bm{\beta}|\sigma^2,\bm{\phi},\bm{y},\bm{X}\sim N(\bm{\beta}_n, \sigma^2{\bf{B}}_n)$, where $\bm{B}_n = (\bm{B}_0^{-1} + \bm{X}^{*\top}\bm{X}^{*})^{-1}$ and $\bm{\beta}_n = \bm{B}_n(\bm{B}_0^{-1}\bm{\beta}_0 + \bm{X}^{*\top}\bm{y}^{*})$. In addition, $\sigma^2|\bm{\beta},\bm{\phi},\bm{y},\bm{X}\sim IG(\alpha_n/2,\delta_n/2)$ where $\alpha_n=\alpha_0+T-p$ and $\delta_n=\delta_0+(\bm{y}^*-\bm{X}^{*}\bm{\beta})^{\top}(\bm{y}^*-\bm{X}^{*}\bm{\beta})+(\bm{\beta}-\bm{\beta}_0)\bm{B}_0^{-1}(\bm{\beta}-\bm{\beta}_0)$. Thus, the previous conditional posterior distributions imply that we can use a Gibbs sampling algorithm to perform inference of these parameters \cite{chib1993bayes}.

We know from Equation \ref{eq1} that $\mu_t=y_t-\bm{x}_t^{\top}\bm{\beta}$, from Equation \ref{eq2} that $\mu_t=\phi_1\mu_{t-1}+\dots+\phi_p\mu_{t-p}+e_t$, $t=p+1,\dots,T$. In matrix notation $\bm{\mu}=\bm{U}\bm{\phi}+\bm{e}$, where $\bm{\mu}$ is a $T-p$ dimensional vector, $\bm{U}$ is a $(T-p)\times p$ matrix whose $t$-th row is $[\mu_{t-1} \ \dots \ \mu_{t-p}]$. Thus, the posterior distribution of $\bm{\phi}|\bm{\beta},\sigma^2,\bm{y},\bm{X}$ is $N(\bm{\phi}_n, \bm{\Phi}_n)\mathbbm{1}[\bm{\phi}\in S_{\bm{\phi}}]$, where $\bm{\Phi}_n=(\bm{\Phi}_0^{-1}+\sigma^{-2}\bm{U}^{\top}\bm{U})$ and $\bm{\phi}_n=\bm{\Phi}_n(\bm{\Phi}_0^{-1}\bm{\phi}_0+\sigma^{-2}\bm{U}^{\top}\bm{\mu})$ (see Exercise 2).

Drawing from the model restricted to stationarity is straightforward: we simply sample from the multivariate normal distribution and discard draws that do not meet the stationarity condition. The proportion of draws that satisfy this restriction represents the conditional probability that the process is stationary.

Observe that the previous setting encompasses the particular relevant case  $y_t\sim AR(p)$, it is just omitting the covariates such that $y_t=\mu_t$. \cite{chib1994bayes} extend the Bayesian inference of linear regression with $AR(p)$ errors to $ARMA(p,q)$ errors.\\

\textbf{Example: Residential water demand I}



Let's assume that the log average quarterly demand of residential water follows an $ARIMA(1,1,1)$ process, that is, $\Delta\log(y_t)\sim ARMA(1,1)$.


Let's use the data set provided by \cite{hassan2018effects} to estimate the demand of residential water in Medellín (Colombia). This a set ....

A nice feature of Bayesian inference is its modular nature. This means we can break down a complex inferential problem into smaller, more manageable parts, this is, a ``divide and conquer" approach. This is possible due to the structure of conditional posterior distributions.  
   
\section{Time varying parameters models}\label{sec83}

Another alternative to handle temporal dependence in a linear regression is to think that the relationship between $y_t$ and $\bm{x}_t$ evolves over times. Then, we can use the \textit{state-space} representation to perform inference in a dynamic linear regression model:

\begin{align*}
	\bm{y}_t&=\bm{x}_t^{\top}\bm{\beta}_t+\mu_t& \text{(Observation equation)}\\
	\bm{\beta}_t&=\bm{\beta}_{t-1}+\bm{w}_t & \text{(States equations)},
\end{align*}
where $\bm{\beta}_0\sim N(\bm{b}_0,\bm{B}_0)$, $\bm{\mu}_t\sim N(\bm{0}, \sigma_t^2)$, $\bm{w}_t\sim N(\bm{0}, \text{diag}\left\{\omega_t^2\right\})$, $\bm{\beta}_0$, $\mu_t$ and $\bm{w}_t$ are independent, $\bm{x}_t$ is a $K$-dimensional vector.

\textbf{Example: Residential water demand III}

Page 167 in \cite{petris2009dynamic} and page 161 in \cite{greenberg2012introduction}.        
     

\section{Stochastic volatility models}\label{sec84}  



\section{Vector Autoregressive models}\label{sec85}

\section{Summary}\label{sec86}

\section{Exercises}\label{sec87}

\begin{enumerate}
	\item Show that in the $AR(2)$ stationary process, $y_t=\mu+\phi_1y_{t-1}+\phi_2y_{t-2}+e_t$, where $e_t\sim N(0,\sigma^2)$, $\mathbb{E}[y_t]=\frac{\mu}{1-\phi_1-\phi_2}$, and $Var[y_t]=\frac{\sigma^2(1-\phi_2)}{1-\phi_2-\phi_1^2-\phi_1^2\phi_2-\phi_2^2+\phi_2^3}$.
	
	\item Show that the posterior distribution of $\bm{\phi}|\bm{\beta},\sigma^2,\bm{y},\bm{X}$ in the model $y_t=\bm{x}_t^{\top}\bm{\beta}+\mu_t$ where $\phi(L)\mu_t=e_t$ and $e_t\stackrel{iid}{\sim}N(0,\sigma^2)$ is $N(\bm{\phi}_n, \bm{\Phi}_n)\mathbbm{1}[\bm{\phi}\in S_{\bm{\phi}}]$, where $\bm{\Phi}_n=(\bm{\Phi}_0^{-1}+\sigma^{-2}\bm{U}^{\top}\bm{U})$, $\bm{\phi}_n=\bm{\Phi}_n(\bm{\Phi}_0^{-1}\bm{\phi}_0+\sigma^{-2}\bm{U}^{\top}\bm{\mu})$, and $S_{\phi}$ is the stationary region of $\bm{\phi}$.
	
	
\end{enumerate}
